---
title: "Week 1 Exercises"
author: "Ray Chandonnet"
date: "October 29, 2022"
output: pdf_document
---

Please complete all exercises below WITHOUT using any libraries/packages.

# Exercise 1
Assign 10 to the variable x. Assign 5 to the variable y. Assign 20 to the variable z.

```{r ex1}
x <- 10
y <- 5
z <- 20
print(x)
print(y)
print(z)
```


# Exercise 2
Show that x is less than z but greater than y.

**Note: your output must be a SINGLE boolean, do not output a boolean for each expression.**

```{r ex2}
Test1 <- x<z & x>y
print(Test1)
```

# Exercise 3
Show that x and y do not equal z. 

**Note: your output must be a SINGLE boolean, do not output a boolean for each expression.**

```{r ex3}
Test2 <- x!=z & y!=z
print(Test2)
```

# Exercise 4
Show that the formula `x + 2y = z`. 

**Note: your output must be a SINGLE boolean**

```{r ex4}
Test3 <- x+2*y == z
print(Test3)
```

# Exercise 5
I have created a vector (Test_vector) of integers for you. Determine if any of x, y, or z are in the vector. 

**Note: your output must be a SINGLE boolean, do not output a boolean for each expression.**

```{r ex5}
Test_vector <- c(1,5,11:22)
x_found <- x %in% Test_vector
y_found <- y %in% Test_vector
z_found <- z %in% Test_vector
Any_found <- x_found | y_found | z_found
print(Any_found)
```

# Exercise 6
Show which value is contained in the test vector. To do this you will need to create an element-wise logical vector using operators. `x == vector`. Once you have done that you will need to use slicing to return all indices that have matches.
**Note: your output should be two integers**

First, I create a vector containing the three values I am searching for (x, y, z)
```{r Ex6Step1}
Searchfor <- c(x,y,z)
```

Now: It seems there are at least two ways to do this.  The first is the way you described above, which is to create a logical vector "Founds" which is the same length as the test vector, and contains booleans indicating whether each value in Test_vector is in the Searchfor vector.  Then you create a new vector that slices the test vector using those booleans.  That code and results are below:

``` {r Ex6Method1}
Founds <- Test_vector %in% Searchfor
Common1 <- Test_vector[Founds]
print(Common1)
```

However, I also discovered thanks to Professor Google that there is a vector function called "Intersect" that returns the same results without the two step process, which might reduce runtime speed if this was a real-life exercise with a very large vector.  I include those results here - I believe that "intersect"" is a base function in R so it doesn't violate the "no libraries / packages" rule of the assignment :)   I would be interested in knowing whether you prefer the approach you recommended or this approach?  Code and Results are below:

``` {r Ex6Method2}
Common2 <- intersect(Searchfor,Test_vector)
print(Common2)
```

One final note I would love your comments on:  My original "programmer's instinct" was to create a Common vector  with no values in it, and then run a For loop, looping through the values I'm searching for and dynamically adding each value to my "Common" if it is found in the test vector.  Is this an inefficient way of doing this?  Or is that in essence what the intersect function is doing "behind the scenes"?  Thanks for indulging me!